MWA tied-array processing III: Microsecond time resolution via a polyphase synthesis filter

### 2020-06-10 ipfb.tex
Put in changes re: updated B0950 plots, and added Keegan to the author list.

\author[McSweeney et al.]{S. J. McSweeney$^1$, S. M. Ord$^2$, D. Kaur$^1$, N. D. R. Bhat$^1$, B. W. Meyers$^{1,2,3}$, S. E. Tremblay, J. Jones$^1$, B. Crosse$^1$
\affil{$^1$International Centre for Radio Astronomy Research (ICRAR), Curtin University, 1 Turner Avenue, Technology Park, Bentley, 6102, W.A., Australia}
\affil{$^2$CSIRO Astronomy and Space Science, PO Box 76, Epping, NSW 1710, Australia}
\affil{$^3$Department of Physics and Astronomy, University of British Columbia, 6224 Agricultural Road, Vancouver, BC V6T 1Z1, Canada}
}

@@

The non-linearity of the back-to-back system induced by Eq. \eqref{eqn:rounded} implies that there is no single impulse response test that can adequately characterise the whole system.
To wit, the results of an impulse response test depend sensitively on at least three factors: the magnitude of the impulse; its position within a tap; and the arbitrary scaling factor and quantisation applied at the end of the test required by the integer-based output formats.
For example, an impulse at tap position $n = 0$ will be perfectly reproduced if its magnitude is chosen to minimise the amount of rounding that takes place during analysis.
One the other hand, an impulse at $n = 64$ can be chosen such that the response is significantly worse than that implied by Fig. \ref{fig:snr}.

@@

\subsubsection{Mitigating rounding errors}

@@

Re-quantisation is beneficial precisely when $\hat{\sigma}_N^2 < \sigma_N^2$, which can be shown by numerical methods to occur when $\sigma_N^2 \lesssim 0.29$.
The MWA test data set presented above has a variance of approximately $6.7$, which implies that the total S/N loss of the back-to-back system would have to be smaller than $-0.05\,$dB before re-quantisation would improve the signal reconstruction.
As Fig. \ref{fig:snr_measured} shows, this is not met for any tap position, so we did not pursue this possibility further.

@@

Observations of three pulsars are presented here to validate the synthesis filter as a useful tool for undertaking high time resolution studies of pulsars with the MWA: MSPs \psrkaurJ{} and \psrbhatJ{}, and the bright, long-period PSR \psrslowB{}.
All three of these observations were taken before the 12-tap least squares synthesis filter had been implemented, and were processed with the mirror filter.
As Fig. \ref{fig:snr_measured} implies, the use of the mirror filter instead of the more optimal least squares filter would result in a further $\sim 0.1\,$dB reduction in S/N.

@@

\psrslowB{} is a bright, long-period ($P = 0.253\,$s) pulsar known to exhibit microstructure \citep{Popov2002,Kuzmin2003}.
A total of eighty seconds (315 pulses) of data ($128 \times 10\,$kHz fine channels) were recorded, beamformed, and subjected to the polyphase synthesis filter.
The mean profile formed from the entire data set is shown in Fig. \ref{fig:0950_80_secs}.
Several individual pulses were sufficiently bright to see substructure.
One such pulse is showcased in Fig. \ref{fig:0950_single_pulse} at three different time resolutions (corresponding to the chosen bin width of each plot).

@@

\begin{figure}[t!]
    \centering
    \includegraphics[width=\columnwidth]{200ms_res.eps} \\[10pt]
    \includegraphics[width=\columnwidth]{50ms_res.eps} \\[10pt]
    \includegraphics[width=\columnwidth]{5ms_res.eps}
    \caption{A bright, single pulse of \psrslowB{}, displayed with full polarisation at time resolutions of $\sim 200\,\mu$s (top), $\sim 50\,\mu$s (middle), and $\sim 5\,\mu$s (bottom). The total time span of the displayed windows are $253\,$ms, $25.3\,$ms, and $2.53\,$ms, respectively.}
    \label{fig:0950_single_pulse}
\end{figure}

Using observations at $102.5\,$MHz, \citet{Kuzmin2003} report microstructure with a characteristic timescale of $\sim 60\mu$s.
Despite the proximity of this timescale to the temporal ``echoes'' induced by the synthesis filter ($50\,\mu$s), we are able to unambiguously identify the $\sim 60\,\mu$s microstructure.
We also suggest that the finest observable structures (Fig. \ref{fig:0950_single_pulse}, bottom plot) might correspond to the $\sim 10\,\mu$s structures reported by \citet{Popov2002} at $1.65\,$GHz.
However, \citet{Kuzmin2003} used only a single linear polarisation feed, and \citet{Popov2002}, only a left-handed circularly polarised feed.
With the benefit of full Stokes polarimetry, we are able to verify that on the smallest time scales, the emission is almost 100\% polarised---mostly linear, but with a small amount of circular polarisation during the brightest part of the pulse.
This is evidence that the errors introduced by the analysis-synthesis filter do not contribute a significant amount of polarisation leakage into neighbouring time bins---in particular, at $50\,\mu$s intervals.

@@

The MWA-VCS telescope system outputs voltage data with a frequency resolution of $10\,$kHz and a time resolution of $100\,\mu$s, which is suitable for many pulsar applications, as demonstrated in several pulsar science papers published to date.
In this paper, we have presented an algorithm to undo the fine channelisation stage (fine PFB) of the VCS pipeline by using a synthesis polyphase filter, implemented as part of our processing software suite, \vcstools{}.
The result is a reconstructed, coarse channelised (pre-PFB) time series, suitable for high-time resolution studies of MSPs and other rapid transient signals.
Although the reconstruction is not perfect, the introduced errors are quantified (no exceeding $-0.65\,$dB for noise-dominated samples).

>>>>

\author[McSweeney et al.]{S. J. McSweeney$^1$, S. M. Ord$^2$, D. Kaur$^1$, N. D. R. Bhat$^1$, B. W. Meyers$^{1,2,3}$, S. E. Tremblay, J. Jones$^1$, B. Crosse$^1$, K. Smith$^1$
\affil{$^1$International Centre for Radio Astronomy Research (ICRAR), Curtin University, 1 Turner Avenue, Technology Park, Bentley, 6102, W.A., Australia}
\affil{$^2$CSIRO Astronomy and Space Science, PO Box 76, Epping, NSW 1710, Australia}
\affil{$^3$Department of Physics and Astronomy, University of British Columbia, 6224 Agricultural Road, Vancouver, BC V6T 1Z1, Canada}
}

@@

The non-linearity of the back-to-back system induced by Eq. \eqref{eqn:rounded} implies that there is no single impulse response test that can adequately characterise the whole system.
To wit, the results of an impulse response test depend sensitively on at least three factors: the magnitude of the impulse; its position within a tap; and the arbitrary scaling factor and quantisation applied at the end of the test required by the integer-based output formats.
For example, an impulse at tap position $n \equiv 0$ (mod $K$) will be perfectly reproduced if its magnitude is such that the amount of rounding that takes place during analysis is minimised.
On the other hand, an impulse at $n \equiv 64$ (mod $K$) can be chosen such that the response is significantly worse than that implied by Fig. \ref{fig:snr}.

@@

\subsubsection*{Mitigation of rounding errors}

@@

Re-quantisation is beneficial precisely when $\hat{\sigma}_N^2 < \sigma_N^2$, which can be shown by numerical methods to occur when $\sigma_N^2 \lesssim 0.29$.
The MWA test data set presented above has a variance of approximately $6.7$, which implies that the total S/N loss of the back-to-back system would have to be smaller than $-0.05\,$dB before re-quantisation would improve the signal reconstruction.
As Fig. \ref{fig:snr_measured} shows, this is not met for any tap position, so we did not pursue this possibility further.
However, systems that are able to achieve better than $\sigma_N^2 \lesssim 0.29$ may do better yet by re-quantising the suitably scaled reconstructed samples.

@@

Observations of three pulsars are presented here to validate the synthesis filter as a useful tool for undertaking high time resolution studies of pulsars with the MWA: MSPs \psrkaurJ{} and \psrbhatJ{}, and the bright, long-period PSR \psrslowB{}.
The first two pulsars were processed with the mirror filter, and \psrslowB{} was processed with the 12-tap least squares filter.
As Fig. \ref{fig:snr_measured} implies, the use of the mirror filter instead of the more optimal least squares filter would result in a further $\sim 0.1\,$dB reduction in S/N.

@@

\psrslowB{} is a bright, long-period ($P = 0.253\,$s) pulsar known to exhibit microstructure \citep{Popov2002,Kuzmin2003}.
A total of eighty seconds (315 pulses) of data ($128 \times 10\,$kHz fine channels) were recorded, beamformed, and subjected to the polyphase synthesis filter.
Across the MWA bandwidth, we used PSRCHIVE's RMFIT to measure an approximate rotation measure (RM) of $1.43 \pm 0.15\,$rad m$^{-2}$, which is consistent with previous measurements of this pulsar's RM \citep[e.g.][]{Noutsos2015}.
The mean profile formed from the one coarse channel is shown in Fig. \ref{fig:0950_80_secs}.
Several individual pulses were sufficiently bright to see substructure.
One such pulse is showcased in Fig. \ref{fig:0950_single_pulse}, where all 24 coarse channels were integrated to maximise the S/N.
To highlight the pulse's substructure, we show it at three different time resolutions (corresponding to the chosen bin width of each plot).

@@

\begin{figure}[t!]
    \centering
    \includegraphics[width=0.7\columnwidth,angle=270]{200us.eps} \\[10pt]
    \includegraphics[width=0.7\columnwidth,angle=270]{50us.eps} \\[10pt]
    \includegraphics[width=0.7\columnwidth,angle=270]{5us.eps}
    \caption{A bright, single pulse of \psrslowB{}, integrated over $24 \times 1.28\,$MHz coarse channels, and displayed with full polarisation at time resolutions of $\sim 200\,\mu$s (top), $\sim 50\,\mu$s (middle), and $\sim 5\,\mu$s (bottom).}
    \label{fig:0950_single_pulse}
\end{figure}

Using observations at $102.5\,$MHz, \citet{Kuzmin2003} report microstructure with a characteristic timescale of $\sim 60\mu$s, which we are able to unambiguously identify in our data.
We suggest that the finest observable structures (Fig. \ref{fig:0950_single_pulse}, bottom plot) might correspond to the $\sim 10\,\mu$s structures reported by \citet{Popov2002} at $1.65\,$GHz.
However, \citet{Kuzmin2003} used only a single linear polarisation feed, and \citet{Popov2002}, only a left-handed circularly polarised feed.
With the benefit of full Stokes polarimetry, we are able to verify that on the smallest time scales, the emission is almost 100\% polarised---mostly linear, but with a small amount of circular polarisation during the brightest part of the pulse.
This is evidence that the errors introduced by the analysis-synthesis filter do not contribute a significant amount of polarisation leakage into neighbouring time bins---in particular, at $50\,\mu$s intervals.

@@

The MWA-VCS telescope system outputs voltage data with a frequency resolution of $10\,$kHz and a time resolution of $100\,\mu$s, which is suitable for many pulsar applications, as demonstrated in several pulsar science papers published to date.
In this paper, we have presented an algorithm to undo the fine channelisation stage (fine PFB) of the VCS pipeline by using a synthesis polyphase filter, implemented as part of our processing software suite, \vcstools{}.
The result is a reconstructed, coarse channelised (pre-PFB) time series, suitable for high-time resolution studies of MSPs and other rapid transient signals.
Although the reconstruction is not perfect, the average error does not exceed $-0.65\,$dB for noise-dominated samples.
